Quasi-universal acceptance of basic science in the US

### 2024-09-19 paper.tex
submission public understanding of science

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\usepackage{csquotes}
\usepackage{placeins}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Quasi-universal acceptance of basic science in the US},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\abstract{%
XX
}

@@

Trust in science is related to many desirable outcomes, from acceptance of anthropogenic climate change (Cologna \& Siegrist, 2020) or vaccination (Lindholt, Jørgensen, Bor, \& Petersen, 2021; Sturgis, Brunton-Smith, \& Jackson, 2021) to following recommendations during COVID (Algan, Cohen, Davoine, Foucault, \& Stantcheva, 2021), which suggests that trust in science was the most important predictor of these behaviors).

@@

Studies of science knowledge provide relevant evidence, as people who answer science knowledge questions correctly presumably trust scientists in this respect. The average level of science knowledge is quite low, even in countries with extensive science education. For example, only 55\% in the US (in 2014) and only 46\% in the EU (in 2005) knew that antibiotics kill only bacteria and not viruses (Science Literacy et al., 2016). However, these results only provide a lower bound on how much people trust basic science, as participants who do not provide the correct answer might accept it when it is pointed out to them.

@@

For Study 2, we restricted our main hypotheses about acceptance to cases in which participants initially provided a wrong answer. However, this meant the more participants had initially provided correct answers, the fewer opportunities they had for accepting correct answers. We provide results on these conditional correlations--for Study 2 and for all other studies--in the ESM\footnote{Only in Study 4 do we find evidence that changing one's mind towards the scientific consensus is associated with (more) trust in science (Studies 1: r = 0.06, p = 0.387; 2: r = 0.16, p = 0.051; 3: r = 0.04, p = 0.619; 4: r = 0.15, p = 0.037) and only in Study 2 evidence that it is associated with (less) conspiracy beliefs and (less) conspiracy thinking (Studies 1: r = -0.14, p = 0.061; 2: r = -0.22, p = 0.006; 3: r = -0.04, p = 0.631; 4: r = -0.05, p = 0.455).}. However, for the analysis presented here, we proceeded as preregistered for all other studies, by reporting unconditional correlations between acceptance and trust in science, or, respectively, conspiracy belief.

@@

After removing failed attention checks, we were left with a total sample size of 782 (194 in Study 1, 6 failed attention checks; 190 in Study 2, 11 failed attention checks; 200 in Study 3, no failed attention checks; 198 in Study 4, 2 failed attention checks) participants in the US via prolific. Details and demographics can be found in the online supplemental material. While samples for Studies 1 and 2 were convenience samples, Studies 3 and 4 were conducted on a sample holding vaccine-skeptic beliefs. Prolific Academic allows selecting participants based on their answers to a range of questions. We picked three of these questions and only recruited participants who met our criteria for each of them:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  ``Please describe your attitudes towards the COVID-19 (Coronavirus) vaccines: {[}For (I feel positively about the vaccines); Against (I feel negatively about the vaccines); Neutral (I don't have strong opinions either way); Prefer not to say''{]}. We selected pariticpants who answered''Against''.
\item
  ``Have you received a coronavirus (COVID-19) vaccination? {[}Yes (at least one dose); No; Prefer not to answer{]}''. We select only people who answered ``No''.
\item
  ``On a scale from 1-7, please rate to what extent you agree with the following statement: I believe that scheduled immunizations are safe for children. {[}1 (totally disagree); 2 (disagree); 3 (somewhat disagree); 4 (neither agree nor disagree); 5 (somewhat agree); 6 (agree); 7 (totally agree); rather not say{]}''. We select only people who answered ``1'', ``2'', or ``3''.
\end{enumerate}

@@

\begin{longtable}[t]{>{\raggedleft\arraybackslash}p{2em}>{\raggedright\arraybackslash}p{22em}>{\raggedright\arraybackslash}p{22em}}
\caption{\label{tab:knowledge}Science knowledge items}\\
\toprule
 & Study 1-3 & Study 4\\
\midrule
1 & Do antibiotics kill viruses as well as bacteria? [Yes, both; No, only viruses; No, only bacteria] & For which disease is the drug bedaquiline, developed in 2007, a treatment? [Tetanus; Tuberculosis; Malaria]\\
2 & Are electrons smaller, larger, or the same size as atoms? [Smaller; Same size; Larger] & What is the maximum speed a proton can attain in the largest particle collider as to 2015? [90\% of the speed of light; 99\% of the speed of light; the speed of light]\\
3 & Have the continents on Earth been moving for millions of years or have they always been where they are now? [They have been moving; They have always been where they are now] & Kepler-452b is an exoplanet revolving around the star Kepler-452. How far away from the star is it, as established by astronomers in 2015? [97 million mi; 1,2 million mi; 1254 million mi]\\
4 & What decides whether a baby is a boy or a girl ? Is it the father's genes, the mother's genes, or both? [The mother's genes; the father's genes; both] & Using bomb-pusle dating with carbon 14, what is the age of the oldest known vertebrate, as established in 2016? [138 years; 205 years;  392 years]\\
5 & Do lasers work by focusing sound waves? [Yes; No] & How many more glial cells are there in the brain in comparison with neurons, as established in 2016? [The same amount; Twice as many; Ten times as many]\\
\addlinespace
6 & How long does it take for Earth to go around the sun: one day, one month, or one year? [One day; One month; One year] & As predicted by the general theory of relativity, how many times would the Earth keep orbiting if the Sun disappeared, as established in 2012? [47 seconds; 8 minutes;  2 hours]\\
7 & Are diamonds made of carbon? [Yes; No] & What is the electric charge of the Higgs Boson, as established in 2012? [1.602176634 x 10-19; 0; 3.2x10-19C]\\
8 & Which travels faster : light or sound? [Light; Sound] & What is the age of the oldest materials formed on Earth, as established in 2020? [Less than 4.6 Ga; Around 4.6 Ga; More than 4.6 Ga]\\
9 & Is common table salt made of calcium carbonate? [Yes; No] & With the best current cloning techniques, what is the average success rate when operated on mice, as of 2010? [2,7\%; 9,4\%; 17,2\%]\\
10 & Is water made of molecules containing one oxygen and two hydrogen atoms? [Yes; No] & What was the strength of the Earth magnetic field 3.7 billion years ago, as discovered this year? [15 microtesla; 30 microtesla; 45 microtesla]\\
\addlinespace
11 & **Where do trees mainly draw the materials with which they create their mass? [Earth; Water; Air] & \\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}\textsuperscript{*} Only used in Study 1}\\
\end{longtable}
\endgroup{}

@@

We used the four-item conspiracy mentality questionnaire (CMQ) (Bruder, Haffke, Neave, Nouripanah, \& Imhoff, 2013) and the single item conspiracy beliefs scale (SICBS) (Lantian, Muller, Nurra, \& Douglas, 2016). Details and comparisons between the scales can be found in the ESM.

@@

In all analyses reported in the main paper, we measure trust in science via a question selected from the Wellcome Global Monitor surveys (Wellcome Global Monitor, 2018, 2020): ``In general, would you say that you trust science a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''. We chose this question as it seemed to be the most general one. In the appendix, we additionally report results for two alternative measures of trust included in our studies: Another from the WGM surveys (``How much do you trust scientists in this country? Do you trust them a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''), and one from the Pew Research Center (``How much confidence do you have in scientists to act in the best interests of the public? {[}1-5; 1 = No confidence at all, 5 = A great deal of confidence{]}''; see e.g. Funk, Johnson, and Hefferon (2019) and used by e.g. Cologna et al. (2024)). We selected these items so that we could compare the answers in our sample to global survey results. We find that all three items are generally highly correlated throughout all studies, and that our results reported here generally replicate when using either of the alternatives measures (with two exceptions: In Study 3, we find no correlation between acceptance and the Pew question; In Study 4, we find a correlation between knowledge and both alternative trust measures, but not with our main measure; see ESM).

@@

The main outcome of interest is acceptance of the scientifically consensual facts presented. Overall, acceptance was very high (aggregating across all studies: 95.1 \%; Study 1: 93 \%; Study 2: 98 \%; Study 3: 98 \%; Study 4: 91 \%). Note that this includes both participants who had previously correctly answered the knowledge question, and participants who changed their mind when presented with the scientific consensus. In Studies 3 and 4, we gave participants a second chance in case they had initially rejected the consensus, which slightly increased acceptance rates in those studies (initial acceptance in Study 3: 96 \%; in Study 4: 86 \%).

@@

Participants in the lowest decile of acceptance still had an average acceptance rate of 67.4 \%. Even participants who considered as ``completely true'' that the earth is flat and who said they do ``not trust science at all'' (N = 3) had an average acceptance rate of 86.7 \%.

These high acceptance rates do not merely reflect science knowledge: participants only correctly answered 65.8 \% of the questions (Study 1: 74 \%; Study 2: 79 \%; Study 3: 75 \%; Study 4: 36 \%) before they were provided with the scientifically consensual answer. Even the lowest decile in science knowledge which, on average, answered correctly only on 20 \% of the questions, had an average acceptance rate of 91.7 \%.

Did participants who had initially provided a wrong answer change their minds towards the scientific consensus? Yes. In most cases (study 1: 76.3 \%; study 2: 92.9 \%; study 3: 95.5 \%; study 4: 89.5 \%), participants readily accepted the scientific consensus after having initially given the wrong answer to a question.

@@

Why did participants reject the scientific consensus? We collected a total of 364 answers (study 2: 35; study 3: 74; study 4: 255) from 167 (study 2: 25; study 3: 47; study 4: 95) different participants to the open-ended questions on why they had rejected the scientific consensus on a particular question. Based on the answers, we created five categories. Table \ref{tab:justifications} summarizes these answers by five categories. All answers can be read in the ESM.

@@

\caption{\label{tab:justifications}Justifications by category for data from studies 2, 3 and 4 combined.}

\begin{tabular}{llll}
\toprule
Category & \multicolumn{1}{c}{N (instances)} & \multicolumn{1}{c}{Share (instances)} & \multicolumn{1}{c}{N (unique participants)}\\
\midrule
Not convinced & 158 & 43.4\% & 70\\
No justification & 103 & 28.3\% & 51\\
Personal convictions & 43 & 11.8\% & 31\\
Mistake & 42 & 11.5\% & 34\\
Religious Beliefs & 18 & 4.9\% & 13\\
\bottomrule
\end{tabular}

Why did participants accept the scientific consensus? In Studies 3 and 4--the vaccine hesitant samples--we had asked participants about cases in which they agreed with the scientific consensus. A total of 320 (Study 3: 122; Study 4: 198) participants answered this question. There were more participants saying they accepted the scientific consensus because they independently verified the fact (Study 3: 47.5\%; Study 4: 47\% ), than participants saying it was because they trust scientists (Study 3: 41.8\%; Study 4: 36.4\%)\footnote{10.7\% in Study 3 and 16.7\% in Study 4 answered with ``other'' and gave an open-ended explanation (see ESM).}. We asked participants who said they verified independently to explain how they did it (answers can be found in the ESM).

In an exploratory analysis, we ran linear regressions to test whether there are differences between participants who said they had trusted science and those who said they had verified the information independently. Perhaps unsurprisingly, those who said they accepted consensus because of trust in scientists reported to trust science more (Study 3: mean = 3; \(\hat{\beta}_{\text{Trust}}\) = 0.19, p = 0.330 on a scale from 1 to 4; Study 4: mean = 2.92; \(\hat{\beta}_{\text{Trust}}\) = 0.46, p \textless{} .001) than those who said they verified independently (Study 3 mean = 2.81; Study 4 mean = 2.45). We did not find a difference regarding acceptance (Study 3: \(\hat{\beta}_{\text{Acceptance}}\) = 0.01, p = 0.523 on a scale from 0 to 1; Study 4: \(\hat{\beta}_{\text{Acceptance}}\) = 0.04, p = 0.116) or regarding beliefs in conspiracy theories (Study 3: \(\hat{\beta}_{\text{BCTI}}\) = -0.16, p = 0.694 on a scale from 1 to 9; Study 4: \(\hat{\beta}_{\text{BCTI}}\) = -0.21, p = 0.322). We also did not find a difference in time spent on the survey in Study 3 (\(\hat{\beta}_{\text{Time}}\) = -0.02, p = 0.985; median = 7.64 mins), but in Study 4, people who said they had accepted the consensus because they they trust scientists tended to spend on average 2 minutes less on the survey (\(\hat{\beta}_{\text{Time}}\) = -2.16, p = 0.018; median = 9.38 mins).\footnote{For these analyses, we excluded outliers who took over 30mins for the survey, which was estimated to take around 10 mins, and for which the median time was 7 minutes. As a result, we excluded 1 participant in Study 3 and 4 in Study 4. Significance levels remain the same when including these outliers.} In Study 4, in which we used facts that participants were unlikely to have encountered before, we tracked whether people clicked on the source links we provided--a behavior that you would expect from people who report verifying facts independently. On average, participants clicked only on 1.36 links (out of 20 possible clicks) and there was no difference between the two groups (\(\hat{\beta}_{\text{Clicks}}\) = -1.09, p = 0.062).

More detailed results addressing all our pre-registered research questions can be found in the ESM.

@@

Overall, the present findings support the motivated rejection of science model (Lewandowsky \& Oberauer, 2016), in which ``people tend to reject findings that threaten their core beliefs or worldview'' (p.~217). A number of participants in our studies rejected basic tenets of science (on evolution, the shape of the earth, etc.). However, they still accepted the vast majority of basic scientific knowledge presented to them. This suggests that these participants had reasons to reject specific scientific knowledge, and that this rejection prompted them to express a lower trust in science when asked general questions on the topic. These participants did not appear to have general grounds for distrusting science, which should have led them to reject most or all of the science knowledge presented to them.

@@

In applied terms, the present results suggest the existence of a vast reservoir of trust in basic science that could be tapped into in science communication. Since people appear so resistant to rejecting basic science, stressing the basic science components of publicly controversial fields, from GMOs to climate change, might help reduce the `consensus gaps' observed in these domains (on climate change, see, e.g. Ranney \& Clark, 2016).

@@

@@

@@

@@

\bibitem[\citeproctext]{ref-committeeonscienceliteracyandpublicperceptionofscienceScienceLiteracyConcepts2016}
Science Literacy, C. on, Public Perception of Science, Board on Science Education, Behavioral, D. of, Sciences, S., Education, \& National Academies of Sciences, Engineering, and Medicine. (2016). \emph{Science Literacy: Concepts, Contexts, and Consequences} (C. E. Snow \& K. A. Dibner, Eds.). Washington, D.C.: National Academies Press. \url{https://doi.org/10.17226/23595}

@@

Did participants who had initially gotten it wrong change their minds towards the scientific consensus? Yes. In most cases (study 1: 76.3 \%; study 2: 92.9 \%; study 3: 95.5 \%; study 4: 89.5 \%), participants readily accepted the scientific consensus after having initially given the wrong answer to a question. In very few cases, participants changed their mind away from the scientific consensus. These participants initially gave the correct response, but rejected the scientific consensus right after, thereby contradicting their own initial response (study 1: 1.6 \%; study 2: 0.5 \%; study 3: 0.7 \%; study 4: 5.7 \%). Figure \ref{fig:acceptance-knowledge} shows acceptance rates conditional on whether the initial answer to the questions was correct or not.

\begin{figure}
\centering
\includegraphics{output/figures/acceptance-knowledge.pdf}
\caption{\label{fig:acceptance-knowledge}Acceptance rates of scientific consensus, based on whether the initial response to the knowledge question was false or true.}
\end{figure}

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Trust in science; conspiracy theories; conspiracy thinking; science knowledge.}
\usepackage{csquotes}
\usepackage{placeins}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Quasi-universal acceptance of basic science in the US},
  pdflang={en-EN},
  pdfkeywords={Trust in science; conspiracy theories; conspiracy thinking; science knowledge.},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\abstract{%
Substantial minorities of the population report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g.~the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1\%), but also in every sub-sample (e.g., no trust in science: 87.3\%; complete endorsement of flat Earth theory: 87.2\%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
}

@@

Trust in science is related to many desirable outcomes, from acceptance of anthropogenic climate change (Cologna \& Siegrist, 2020) or vaccination (Lindholt, Jørgensen, Bor, \& Petersen, 2021; Sturgis, Brunton-Smith, \& Jackson, 2021) to following recommendations during COVID (Algan, Cohen, Davoine, Foucault, \& Stantcheva, 2021, which suggests that trust in science was the most important predictor of these behaviors).

@@

Studies of science knowledge provide relevant evidence, as people who answer science knowledge questions correctly presumably trust scientists in this respect. The average level of science knowledge is quite low, even in countries with extensive science education. For example, only 55\% in the US (in 2014) and only 46\% in the EU (in 2005) knew that antibiotics kill only bacteria and not viruses (National Academies of Sciences, Engineering, and Medicine, 2016). However, these results only provide a lower bound on how much people trust basic science, as participants who do not provide the correct answer might accept it when it is pointed out to them.

@@

For Study 2, we restricted our main hypotheses about acceptance to cases in which participants initially provided a wrong answer. However, this meant the more participants had initially provided correct answers, the fewer opportunities they had for accepting correct answers. We provide results on these conditional correlations--for Study 2 and for all other studies--in the ESM, section \ref{exp1} to \ref{exp4} \footnote{Only in Study 4 do we find evidence that changing one's mind towards the scientific consensus is associated with (more) trust in science (Studies 1: r = 0.06, p = 0.387; 2: r = 0.16, p = 0.051; 3: r = 0.04, p = 0.619; 4: r = 0.15, p = 0.037) and only in Study 2 evidence that it is associated with (less) conspiracy beliefs and (less) conspiracy thinking (Studies 1: r = -0.14, p = 0.061; 2: r = -0.22, p = 0.006; 3: r = -0.04, p = 0.631; 4: r = -0.05, p = 0.455).}. However, for the analysis presented here, we proceeded as preregistered for all other studies, by reporting unconditional correlations between acceptance and trust in science, or, respectively, conspiracy belief.

@@

After removing failed attention checks, the total sample size was 782 (194 in Study 1, six failed attention checks; 190 in Study 2, 11 failed attention checks; 200 in Study 3, no failed attention checks; 198 in Study 4, two failed attention checks) participants in the US, recruited through Prolific. Details and demographics can be found in the online supplemental material. While samples for Studies 1 and 2 were convenience samples, Studies 3 and 4 were conducted on a sample holding vaccine-skeptic beliefs. Prolific allows selecting participants based on their answers to a range of questions. We picked three of these questions and only recruited participants who met our criteria for each of them:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  ``Please describe your attitudes towards the COVID-19 (Coronavirus) vaccines: {[}For (I feel positively about the vaccines); Against (I feel negatively about the vaccines); Neutral (I don't have strong opinions either way); Prefer not to say{]}''. We selected participants who answered ``Against''.
\item
  ``Have you received a coronavirus (COVID-19) vaccination? {[}Yes (at least one dose); No; Prefer not to answer{]}''. We select only people who answered ``No''.
\item
  ``On a scale from 1-7, please rate to what extent you agree with the following statement: I believe that scheduled immunizations are safe for children. {[}1 (totally disagree); 2 (disagree); 3 (somewhat disagree); 4 (neither agree nor disagree); 5 (somewhat agree); 6 (agree); 7 (totally agree); rather not say{]}''. We select only people who answered ``1'', ``2'', or ``3''.
\end{enumerate}

@@

\begin{longtable}[t]{>{\raggedleft\arraybackslash}p{2em}>{\raggedright\arraybackslash}p{22em}>{\raggedright\arraybackslash}p{22em}}
\caption{\label{tab:knowledge}Science knowledge items}\\
\toprule
 & Study 1-3 & Study 4\\
\midrule
1 & Do antibiotics kill viruses as well as bacteria? [Yes, both; No, only viruses; No, only bacteria] & For which disease is the drug bedaquiline, developed in 2007, a treatment? [Tetanus; Tuberculosis; Malaria]\\
2 & Are electrons smaller, larger, or the same size as atoms? [Smaller; Same size; Larger] & What is the maximum speed a proton can attain in the largest particle collider as to 2015? [90\% of the speed of light; 99\% of the speed of light; the speed of light]\\
3 & Have the continents on Earth been moving for millions of years or have they always been where they are now? [They have been moving; They have always been where they are now] & Kepler-452b is an exoplanet revolving around the star Kepler-452. How far away from the star is it, as established by astronomers in 2015? [97 million mi; 1,2 million mi; 1254 million mi]\\
4 & What decides whether a baby is a boy or a girl ? Is it the father's genes, the mother's genes, or both? [The mother's genes; the father's genes; both] & Using bomb-pusle dating with carbon 14, what is the age of the oldest known vertebrate, as established in 2016? [138 years; 205 years;  392 years]\\
5 & Do lasers work by focusing sound waves? [Yes; No] & How many more glial cells are there in the brain in comparison with neurons, as established in 2016? [The same amount; Twice as many; Ten times as many]\\
\addlinespace
6 & How long does it take for Earth to go around the sun: one day, one month, or one year? [One day; One month; One year] & As predicted by the general theory of relativity, how many times would the Earth keep orbiting if the Sun disappeared, as established in 2012? [47 seconds; 8 minutes;  2 hours]\\
7 & Are diamonds made of carbon? [Yes; No] & What is the electric charge of the Higgs Boson, as established in 2012? [1.602176634 x 10-19; 0; 3.2x10-19C]\\
8 & Which travels faster : light or sound? [Light; Sound] & What is the age of the oldest materials formed on Earth, as established in 2020? [Less than 4.6 Ga; Around 4.6 Ga; More than 4.6 Ga]\\
9 & Is common table salt made of calcium carbonate? [Yes; No] & With the best current cloning techniques, what is the average success rate when operated on mice, as of 2010? [2,7\%; 9,4\%; 17,2\%]\\
10 & Is water made of molecules containing one oxygen and two hydrogen atoms? [Yes; No] & What was the strength of the Earth magnetic field 3.7 billion years ago, as discovered this year? [15 microtesla; 30 microtesla; 45 microtesla]\\
\addlinespace
11 & *Where do trees mainly draw the materials with which they create their mass? [Earth; Water; Air] & \\
\bottomrule
\multicolumn{3}{l}{\rule{0pt}{1em}\textsuperscript{*} Only used in Study 1}\\
\end{longtable}
\endgroup{}

@@

For all results presented here, we used the four-item conspiracy mentality questionnaire (CMQ) (Bruder, Haffke, Neave, Nouripanah, \& Imhoff, 2013). We also assessed the single item conspiracy beliefs scale (SICBS) (Lantian, Muller, Nurra, \& Douglas, 2016). Details and comparisons between the scales can be found in the ESM, sections \ref{exp1} to \ref{exp4}.

@@

In all analyses reported in the main paper, we measure trust in science via a question selected from the Wellcome Global Monitor surveys (Wellcome Global Monitor, 2018, 2020): ``In general, would you say that you trust science a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''. We chose this question as it seemed to be the most general one. In the ESM (sections \ref{exp1} to \ref{exp4}), we additionally report results for two alternative measures of trust included in our studies: Another from the WGM surveys (``How much do you trust scientists in this country? Do you trust them a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''), and one from the Pew Research Center (e.g. Funk, Johnson, \& Hefferon, 2019) (``How much confidence do you have in scientists to act in the best interests of the public? {[}1-5; 1 = No confidence at all, 5 = A great deal of confidence{]}''), the latter having been used in a recent international study on trust in science (Cologna et al., 2024). We selected these items so that we could compare the answers in our sample to global survey results. We find that all three items are generally highly correlated throughout all studies, and that our results reported here generally replicate when using either of the alternatives measures (with two exceptions: In Study 3, we find no correlation between acceptance and the Pew question; In Study 4, we find a correlation between knowledge and both alternative trust measures, but not with our main measure; see ESM, sections \ref{exp3} and \ref{exp4}).

@@

The main outcome of interest is acceptance of the scientifically consensual facts presented. Overall, acceptance was very high (aggregating across all studies: 95.1 \%; Studies 1: 93 \%; 2: 98 \%; 3: 98 \%; 4: 91 \%). Note that this includes both participants who had previously correctly answered the knowledge question, and participants who changed their mind when presented with the scientific consensus. In Studies 3 and 4, we gave participants a second chance in case they had initially rejected the consensus, which slightly increased acceptance rates in those studies (initial acceptance in Studies 3: 96 \%; 4: 86 \%).

@@

Participants in the lowest decile of acceptance still had an average acceptance rate of 67.4 \%. Even the three participants who considered as ``completely true'' that the earth is flat and who said they do ``not trust science at all'' had an average acceptance rate of 86.7 \%.

These high acceptance rates do not merely reflect science knowledge: participants only correctly answered 65.8 \% of the questions (Studies 1: 74 \%; 2: 79 \%; 3: 75 \%; 4: 36 \%) before they were provided with the scientifically consensual answer. Even the lowest decile in science knowledge, which on average answered correctly only on 20 \% of the questions, had an average acceptance rate of 91.7 \%.

Did participants who had initially provided a wrong answer change their minds towards the scientific consensus? Yes. In most cases (Studies 1: 76.3 \%; 2: 92.9 \%; 3: 95.5 \%; 4: 89.5 \%), participants readily accepted the scientific consensus after having initially given the wrong answer to a question.

@@

Why did participants reject the scientific consensus? We collected a total of 364 answers (Studies 2: 35; 3: 74; 4: 255) from 167 (Studies 2: 25; 3: 47; 4: 95) participants to the open-ended questions on why they had rejected the scientific consensus on a particular question. Based on the answers, we created five categories (Table \ref{tab:justifications}). All answers can be read in the ESM, sections \ref{exp2} to \ref{exp4}.

@@

\caption{\label{tab:justifications}Justifications for rejecting the scientific consensus by category, Studies 2, 3 and 4 combined.}

\begin{tabular}{llll}
\toprule
Category & \multicolumn{1}{c}{N (instances)} & \multicolumn{1}{c}{Share (instances)} & \multicolumn{1}{c}{N (participans)*}\\
\midrule
Not convinced & 158 & 43.4\% & 70\\
No justification & 103 & 28.3\% & 51\\
Personal convictions & 43 & 11.8\% & 31\\
Mistake & 42 & 11.5\% & 34\\
Religious Beliefs & 18 & 4.9\% & 13\\
\bottomrule
\addlinespace
\end{tabular}

\begin{tablenotes}[para]
\normalsize{\textit{Note.} *Participans with at least one answer in that category}
\end{tablenotes}

Why did participants say they accept the scientific consensus? In Studies 3 and 4--the vaccine hesitant samples--we had asked participants about cases in which they agreed with the scientific consensus. A total of 320 (Studies 3: 122; 4: 198) participants answered this question. There were more participants saying they accepted the scientific consensus because they independently verified the fact (Studies 3: 47.5\%; 4: 47\% ), than participants saying it was because they trust scientists (Studies 3: 41.8\%; 4: 36.4\%)\footnote{10.7\% in Study 3 and 16.7\% in Study 4 answered with ``other'' and gave an open-ended explanation (see ESM, sections \ref{exp3} and \ref{exp4}).}. Answers to a question about how they had done so can be found in the ESM, sections \ref{exp3} and \ref{exp4}.

In an exploratory analysis, we ran linear regressions to test whether there are differences between participants who said they had trusted science and those who said they had verified the information independently. Participants who said they accepted the consensus because of trust in scientists reported trusting science more (Studies 3: mean = 3; \(\hat{\beta}_{\text{Trust}}\) = 0.19, p = 0.330 on a scale from 1 to 4; 4: mean = 2.92; \(\hat{\beta}_{\text{Trust}}\) = 0.46, p \textless{} .001) than those who said they verified independently (Studies 3 mean = 2.81; 4: mean = 2.45). We did not find a difference regarding acceptance (Studies 3: \(\hat{\beta}_{\text{Acceptance}}\) = 0.01, p = 0.523 on a scale from 0 to 1; 4: \(\hat{\beta}_{\text{Acceptance}}\) = 0.04, p = 0.116) or regarding beliefs in conspiracy theories (Studies 3: \(\hat{\beta}_{\text{BCTI}}\) = -0.16, p = 0.694 on a scale from 1 to 9; 4: \(\hat{\beta}_{\text{BCTI}}\) = -0.21, p = 0.322). We also did not find a difference in time spent on the survey in Study 3 (\(\hat{\beta}_{\text{Time}}\) = -0.02, p = 0.985; median = 7.64 mins), but in Study 4, people who said they had accepted the consensus because they they trust scientists tended to spend on average two minutes less on the survey (\(\hat{\beta}_{\text{Time}}\) = -2.16, p = 0.018; median = 9.38 mins).\footnote{For these analyses, we excluded outliers who took over 30 mins for the survey, which was estimated to take around 10 mins, and for which the median time was seven minutes. As a result, we excluded one participant in Study 3 and four in Study 4. Significance levels are not affected by these exclusions.} In Study 4, in which we used facts that participants were unlikely to have encountered before, we tracked whether people clicked on the source links we provided--a behavior that you would expect from people who report verifying facts independently. On average, participants clicked only on 1.36 links (out of 20 possible clicks) and there was no difference between the two groups (\(\hat{\beta}_{\text{Clicks}}\) = -1.09, p = 0.062).

More detailed results addressing all our pre-registered research questions can be found in the ESM, sections \ref{exp1} to \ref{exp4}.

@@

Overall, the present findings support the motivated rejection of science model (Lewandowsky \& Oberauer, 2016), in which ``people tend to reject findings that threaten their core beliefs or worldview'' (p.~217). A number of participants in our studies rejected basic tenets of science (on evolution, the shape of the earth, etc.). However, they still accepted the vast majority of basic scientific knowledge presented to them. This suggests that these participants had reasons to reject specific scientific knowledge, and that this rejection prompted them to express a lower trust in science when asked general questions on the topic. These participants did not appear to have general grounds for distrusting science, which should have led them to reject most or all of the science knowledge presented to them. These results are in line with several recent studies in which conspiracy theorists are not less susceptible to social influence in general (Altay, Nera, Ejaz, Schöpfer, \& Tomas, 2023; Pummerer, Fock, Winter, \& Sassenberg, n.d.).

@@

In applied terms, the present results suggest the existence of a vast reservoir of trust in basic science that can be tapped into in science communication. Since people appear so resistant to rejecting basic science, stressing the basic science components of publicly controversial fields, from GMOs to climate change, might help reduce the `consensus gaps' observed in these domains (on climate change, see, e.g. Ranney \& Clark, 2016).

@@

\bibitem[\citeproctext]{ref-altayConspiracyBelieversClaim2023}
Altay, S., Nera, K., Ejaz, W., Schöpfer, C., \& Tomas, F. (2023). Conspiracy believers claim to be free thinkers but (Under)Use advice like everyone else. \emph{British Journal of Social Psychology}, \emph{62}(4), 1782--1797. \url{https://doi.org/10.1111/bjso.12655}

@@

\bibitem[\citeproctext]{ref-nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016}
National Academies of Sciences, Engineering, and Medicine. (2016). \emph{Science Literacy: Concepts, Contexts, and Consequences} (C. E. Snow \& K. A. Dibner, Eds.). Washington, D.C.: National Academies Press. \url{https://doi.org/10.17226/23595}

@@

\bibitem[\citeproctext]{ref-pummererConspiracyBeliefsMajority}
Pummerer, L., Fock, L., Winter, K., \& Sassenberg, K. (n.d.). Conspiracy beliefs and majority influence. \emph{The Journal of Social Psychology}, \emph{0}(0), 1--16. \url{https://doi.org/10.1080/00224545.2024.2397491}

@@

@@

Did participants who gave a wrong answer change their minds towards the scientific consensus? In most cases (Studies 1: 76.3 \%; 2: 92.9 \%; 3: 95.5 \%; 4: 89.5 \%), participants readily accepted the scientific consensus after having initially given the wrong answer to a question. In very few cases, participants changed their mind away from the scientific consensus. These participants initially gave the correct response, but rejected the scientific consensus right after, thereby contradicting their own initial response (Studies 1: 1.6 \%; 2: 0.5 \%; 3: 0.7 \%; 4: 5.7 \%). Figure \ref{fig:acceptance-knowledge} shows acceptance rates conditional on whether the initial answer to the questions was correct or not.

\begin{figure}
\centering
\includegraphics{output/figures/acceptance-knowledge.pdf}
\caption{\label{fig:acceptance-knowledge}Acceptance/Rejection rate of scientific consensus, based on whether the initial response to the knowledge question was false or true.}
\end{figure}

\section{Trust in science}\label{trust}

Figure \ref{fig:trust-overview} shows the distributions of our main trust in science measure across the four studies.

\begin{figure}
\centering
\includegraphics{output/figures/trust-overview.pdf}
\caption{\label{fig:trust-overview}Distributions of trust in science.}
\end{figure}

\section{Conspiracy Thinking}\label{conspiracy-thinking}

Figure \ref{fig:conspiracy-thinking-overview} shows the distributions of conspiracy thinking across the four studies.

\begin{figure}
\centering
\includegraphics{output/figures/conspiracy-thinking-overview.pdf}
\caption{\label{fig:conspiracy-thinking-overview}Distributions of conspiracy thinking.}
\end{figure}

\section{Conspiracy Belief}\label{conspiracy-belief}

Figure \ref{fig:conspiracy-belief-overview} shows the average belief in each of the conspiracy theories across the four studies.

\begin{figure}
\centering
\includegraphics{output/figures/conspiracy-belief-overview.pdf}
\caption{\label{fig:conspiracy-belief-overview}Average belief in conspiracy theories.}
\end{figure}

\section{Science knowledge}\label{knowledge}

Figure \ref{fig:knowledge-overview} shows the distributions of basic science knowledge across the four studies.

\begin{figure}
\centering
\includegraphics{output/figures/knowledge-overview.pdf}
\caption{\label{fig:knowledge-overview}Distributions of science knowledge.}
\end{figure}

\section{Acceptance and knowledge by question}\label{by-question-variation}
